Distributional Trees for Circular Data

### 2019-02-01 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Include Reto's input (application)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2422 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 447dfddb4da7617048853054c6ca37a7168c18e4

\begin{figure}
\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
\label{fig:tree}
\end{figure}

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision making processes and risk management,
e.g., in air traffic management and renewable energy production. As a suitable
and useful application of circular regression trees, we therefore forecast the
wind direction at the airport of Innsbruck. Innsbruck lies at the bottom of the
Inn Valley, a deep mountain valley in the Eastern European Alps orientated
parallel to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is
bordered in the North with a approx. 2300\,m high mountain ridge and opens in
the South into the crest-perpendicular Wipp Valley. As potential split
variables we employ several 1-hourly lagged observed weather variables (e.g.,
wind direction, wind speed, relative humidity, pressure gradients along the
valley) and derived spatial and temporal difference of these lagged
observations.

As an illustration of the tree based wind direction forecasts at Innsbruck airport, a
BIC based post-pruned regression tree of depth three is shown in
Fig.~\ref{fig:tree}. Three distinct wind regimes are fitted within the final
nodes: Firstly, up-valley winds blowing from the valley mouth towards the upper
valley (from east to west, as predicted in the final nodes 4 and 5). Secondly,
downslope winds blowing across the Alpine crest along the Wipp Valley towards
Innsbruck (from south-east to north-west, as predicted in nodes 7 and 8).
Thirdly, down-valley winds blowing in the direction of the valley mouth (from
west to east, as predicted in nodes 12, 14, and 15). The final node 11 with a
concentration parameter $\kappa$ of only $0.84$ captures the cases with rather low
wind speeds and typically no clear wind direction or distinct wind regime.
In general, winds typically blow from areas of high pressure to areas of low
pressure, and a strong pressure gradient leads to large wind speeds and a stable
wind direction. Therefore, the splits in the lagged wind direction perform a
clustering into the different wind regimes and the further splits in pressure
gradients or wind speed divide the predictions according to the consistency of
the lagged wind direction. This can be seen in the final nodes with strong
pressure gradients or high wind speeds where most training cases are classified
correctly and therefore the concentration parameter as an indicator of the
certainty of the forecast is rather large.

>>>>

%\begin{figure}
%\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
%\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
%\label{fig:tree}
%\end{figure}

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision making processes and risk management,
e.g., in air traffic management or renewable energy production. This study
presents an application of circular regression trees  on hourly direction
forecasts at Innsbruck airport.  Innsbruck lies at the bottom of the Inn
Valley, a deep mountain valley in the Eastern European Alps orientated parallel
to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered
in the North with a approx. 2300\,m high mountain ridge and opens to the South
into the crest-perpendicular Wipp Valley. As potential split variables we
employ several 1-hourly lagged observed weather variables (e.g., wind
direction, wind speed, relative humidity, pressure gradients along the valley)
and derived spatial and temporal difference of these lagged observations.

As an illustration of the tree based wind direction forecasts at Innsbruck
airport, a BIC based post-pruned regression tree of depth three is shown in
Fig.~\ref{fig:tree}. Three distinct wind regimes are fitted within the final
nodes: Firstly, up-valley winds blowing from the valley mouth towards the upper
valley (from east to west, nodes 4 and 5). Secondly, downslope winds blowing
across the Alpine crest along the Wipp Valley towards Innsbruck (from
south-east to north-west, nodes 7 and 8). Thirdly, down-valley winds blowing
in the direction of the valley mouth (from west to east, nodes 12, 14, and 15).
The final node 11 with a concentration parameter $\kappa$ of only $0.84$
captures the cases with rather low wind speeds which typically don't show a
clear wind direction or distinct wind regime. In general, winds typically blow
from areas of high pressure to areas of low pressure where strong pressure
gradients yield high wind speeds and a rather well-defined wind direction.
Thus, the splits in the lagged wind direction perform a clustering into the
different wind regimes and the additional splits in pressure gradients or wind
speed divide the predictions according to the expected consistency of the wind
direction. This can be seen in the final nodes with strong pressure gradients
or high wind speeds where most training cases are classified correctly and
therefore the concentration parameter as an indicator of the certainty of the
forecast is rather large.
